---
title: "run_analysis"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown for run_analysis

This markdown explains the steps taken by the function run_analysis(dir) to create a tidy dataset from the "UCI HAR Dataset"

The dir is of type character that can be passed into run_analysis to define where in the current working directory to find the txt files.

```{r, echo=FALSE}

library(dplyr)
library(data.table)
library(tidyr)

```

Set files directory "UCI HAR Dataset" into wd

```{r}

wd <- paste0(getwd(),"/","UCI HAR Dataset")

```


Put activity_labels.txt in activity_labels
and set column names to be "activityid" and "activity"

```{r}

  # get activity labels and sets column names
  activity_labels <- read.table(paste0(wd,"/activity_labels.txt")) %>%
    setnames(c("activityId","activity"))
  head(activity_labels)

```

set features.txt list into features

```{r}

  # get features
  features <- read.table(paste0(wd,"/features.txt"))
  head(features)

```

Filter features to only include mean() and std() into mean_std.features

```{r}
  # filter features for only mean() and std()
  mean_std.features <- features %>%
    filter(grepl("mean\\(\\)|std\\(\\)",V2))
  head(mean_std.features)

```

Set subject_test.txt data to subject_test and set the column name as "subject"

```{r}

  # get subject_test and set column name
  subject_test <- read.table(paste0(wd,"/test/subject_test.txt")) %>%
    setnames("subject")
  head(subject_test)

```

set y_test.txt data to y_test and set column name as "activityId"

```{r}

  # get y_test and set column name
  y_test <- read.table(paste0(wd,"/test/y_test.txt")) %>%
    setnames("activityId")
  head(y_test)

```

set X_test.txt data to X_test

```{r}

  # get X_test
  X_test <- read.table(paste0(wd,"/test/X_test.txt"))
  head(X_test)

```

get subject_train.txt data into subject_train

```{r}

  # get subject_train and set column name
  subject_train <- read.table(paste0(wd,"/train/subject_train.txt")) %>%
    setnames("subject")
  head(subject_train)

```

set y_train.txt data to y_train and set column name as "activityId"

```{r}

  # get y_train and set column name
  y_train <- read.table(paste0(wd,"/train/y_train.txt")) %>%
    setnames("activityId")
  head(y_train)

```

set X_train.txt data to X_train

```{r}

  # get X_train
  X_train <- read.table(paste0(wd,"/train/X_train.txt"))
  head(X_train)

```

row bind activities of train and test (train first) and put in activity

```{r}

  # row bind activities of train and test (train first)
  activity <- rbind(y_train,y_test)
  head(activity)

```

row bind subjects of train and test (train first) and put in subject

```{r}

  # row bind subjects of train and test (train first)
  subject <- rbind(subject_train,subject_test)
  head(subject)

```

Create the tidy dataset test_and_train:
1. row bind the X_train and X_test
2. column bind the subject to the result of 1
3. column bind the activity to the result of 2
4. convert the columns of feature types to one column (gather on feature)
5. create featureid column from the feature column (remove the "V")
6. merge with the mean_std.features to get filter for mean() and std() features and get the feature names
6. merge the activity_labels to get the labels
7. select and name the relevant columns: subject, activity, V2, result and set column name V2 to variable

```{r}

  # Merges the training and the test sets to create one data set.
  test_and_train <- rbind(X_train, X_test) %>%
    # adds subject column to data set
    cbind(subject) %>%
    # adds activity column to data set
    cbind(activity) %>%
    # Convert columns of observations to one column
    gather(feature, result, -c(subject, activityId)) %>%
    # Get feature id
    mutate(featureid = as.integer(gsub("[^0-9]","", feature))) %>%
    # Get mean and std feature descriptions
    merge(mean_std.features, by.x = "featureid", by.y = "V1") %>%
    # Uses descriptive activity names to name the activities in the data set
    merge(activity_labels) %>%
    # separate the feature description into measurement, type and axis
    separate('V2', c("measurement", "measurementType", "axis"), sep="-", extra="drop") %>%
    # Extracts only the measurements on the mean and standard deviation for each measurement.
    select(subject, activity, measurement, measurementType, axis, result) %>%
    # Order
    arrange(subject, activity, measurement, measurementType, axis)
  head(test_and_train)

```

Aggregate the test_and_train to get the mean of the results by grouping on subject, activity and variable

```{r}

  # aggregate numeric columns by activity and subject, removing NA values
  aggdata <-test_and_train %>%
    # group by subject, activity measurement, measurementType and axis
    group_by(subject, activity, measurement, measurementType, axis) %>%
    # summarize by the mean of the result
    summarize(mean = mean(result))
  head(aggdata)


```

